---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "output/README-",
  out.width = "100%"
)
```

# CytoProfile

<!-- badges: start -->
<!-- badges: end -->

The goal of CytoProfile is to conduct quality control using biological
meaningful cutoff on raw measured values of cytokines. Specifically, test on 
distributional symmetry to suggest the adopt of transformation. 
Conduct exploratory analysis including summary statistics, generate enriched 
barplots, and boxplots. Further, conduct univariate analysis and 
multivariate analysis for advance analysis. 

## Installation

Before installation of the CytoProfile package, make sure to install BiocManager and mix0mics packages using: 

```r
## install BiocManager
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager") 
## install mixOmics 
BiocManager::install('mixOmics')
```


You can install the development version of CytoProfile from [GitHub](https://github.com/saraswatsh/CytoProfile) with:

``` r
# install.packages("devtools")
devtools::install_github("saraswatsh/CytoProfile")
```

## Example

Below are examples of using the functions provided in CytoProfile. Any saved files whether PDF or PNG is available in the output folder. 

## 1. Data Loading and set up

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Loading all packages required
library(tidyverse) # Used to load dplyr, ggplot2, and tidyr
library(knitr) # Required for setting the working directory
library(mixOmics) # Required for sPLS-DA
library(pROC) # Required for ROC curve in XGBoost and Random Forest
library(moments) # Required for skewness and kurtosis
library(CytoProfile)
# Loading in data
data("cytodata")
data.df <- cytodata

## Setting working directory to output folder to save the PDF files. 
opts_knit$set(root.dir = "E:/Desktop/RA/R Package/CytoProfile/output")
```

## 2. Exploratory Data Analysis

### Boxplots

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating boxplots to check for outliers for raw values
cyt.bp(data.df[,-c(1:4)], Title = "Boxplot.byCytokine.Raw.pdf") # We are removing the first 4 columns as we only want the continuous variables. 

# Generating boxplots to check for outliers for log2 values
cyt.bp(log2(data.df[,-c(1:4)]), Title = "Boxplot.byCytokine.log2.pdf") # Make sure to use log2 to transform the cytokines and same reason as above for removing initial columns

```

### Group-Specific Boxplots

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Second function for generating boxplots
# Raw values
cyt.bp2(data.df[,-c(1,4)], Title = "Boxplot.byGroupandTreatment.raw.pdf", scale = NULL)
# Log-2 transformation
cyt.bp2(data.df[,-c(1,4)], Title = "Boxplot.byGroupandTreatment.log2.pdf", scale = "log2")

```

## 3. Skewness and Kurtosis 

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating histograms for skewness and kurtosis based on raw values and log2 transformation
cyt.skku(data.df[,-c(1,4)], Title = "Skew and Kurtosis.pdf")
```

## 4. Error Bar Plots

### Basic Error Bar Plot

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating Error Bar Plot
cytokine.mat <- cytodata[, -c(1:4)] # Extracting all cytokines to be stored in one object
cytokineNames <- colnames(cytokine.mat) # Extracting the cytokine names
nCytokine <- length(cytokineNames) # Obtaining the total number of cytokines
results <- cyt.skku(cytodata[,-c(1,4)], printResLog = TRUE) # Extracting values
pdf( "barErrorPlot.pdf" )
par(mfrow=c(2,2), mar=c(8.1,  4.1, 4.1, 2.1) )
for( k in 1:nCytokine ) {
  center.df <- data.frame( "name"=rownames(results[,,k]), results[,,k] )
  cyt.errbp(center.df, pLab=FALSE, esLab=FALSE, classSymbol=TRUE,
               ylab="Concentration in log2 scale",  main=cytokineNames[k] )
}
dev.off()
```

### Enriched Error Bar Plot with p-values and Effect Sizes

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating Error Bar Plot enriched with p-value and effect size 
data.df <- cytodata[,-1]
cyt.mat <- log2(data.df[,-c(1:3)])
data.df1 <- data.frame(data.df[,c(1:3)], cyt.mat)
cytokineNames <- colnames(cyt.mat)
nCytokine <- length(cytokineNames)
condt <- !is.na(cyt.mat) & cyt.mat >0
Cutoff <- min(cyt.mat[condt], na.rm=TRUE)/10
# Creating a matrix for p-values from anova tests
p.aov.mat <- matrix(NA, nrow=nCytokine, ncol=3)
# Changing column names
dimnames(p.aov.mat) <- list( cytokineNames, c("Group", "Treatment", "Interaction") )
# Matrix to extract p-values from Tukey group comparison
p.groupComp.mat <- matrix(NA, nrow=nCytokine, ncol=3)
# Changing column names
dimnames(p.groupComp.mat) <- list( cytokineNames, c("2-1", "3-1", "3-2") )
# Matrix for SSMD same size as other matrices
ssmd.groupComp.stm.mat <- mD.groupComp.stm.mat <- p.groupComp.stm.mat <- p.groupComp.mat

for( i in 1:nCytokine ) {
   #i <- 1 # i<-2
  Cytokine <- (cyt.mat[,i]+Cutoff)
  cytokine.aov <- aov( Cytokine ~ Group * Treatment, data=data.df)
  aov.table <- summary(cytokine.aov)[[1]]
  p.aov.mat[i,] <- aov.table[1:3,5]
  p.groupComp.mat[i,] <- TukeyHSD(cytokine.aov)$Group[1:3,4]
  p.groupComp.stm.mat[i,] <- TukeyHSD(cytokine.aov)$`Group:Treatment`[c(1:3),4]
  mD.groupComp.stm.mat[i,] <- TukeyHSD(cytokine.aov)$`Group:Treatment`[c(1:3),1]
  ssmd.groupComp.stm.mat[i,]<-mD.groupComp.stm.mat[i,]/sqrt(2*aov.table["Residuals","Mean Sq"])
}

results <- cyt.skku(cytodata[,-c(1,4)], printResLog = TRUE)
pdf( "barErrorPlot.enriched.pdf" )
par(mfrow=c(2,3), mar=c(8.1,  4.1, 4.1, 2.1) )
for( k in 1:nCytokine ) {
  #k <- 1
  result.mat <- results[1:9,,k]
  center.df <-
    data.frame( "name"=rownames(result.mat), result.mat[, c("center", "spread")],
                "p.value"= c(1,p.groupComp.stm.mat[k,1:2]),
                "effect.size"=c(0,ssmd.groupComp.stm.mat[k,1:2])
    )
  cyt.errbp(center.df, pLab=TRUE, esLab=TRUE, classSymbol=TRUE,
               ylab="Concentration in log2 scale", main=cytokineNames[k])
}
dev.off()
```

## 5. Univariate Analysis

### Two Sample T-test and Mann Whitney U Test

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Performing Two Sample T-test and Mann Whitney U Test
data.df <- cytodata[,-c(1,4)]
data.df <- filter(data.df, Group != "ND", Treatment != "Unstimulated")
# Two sample T-test
cyt.ttests(data.df, scale = "log2")
# Mann Whitney U Test
cyt.ttests(data.df)

```

### ANOVA Comparisons Test

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Performing ANOVA comparisons test for univariate analysis
cyt.anova(data.df[,c(1:2,5:6)]) # This only considers 2 cytokines for this example only
```

## 6. Multivariate Analysis

### Partial Least Squares Discriminant Analysis (PLS-DA) 

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# In this code, we will have background predict to be turned on to see the classification areas and 
# we will also print out the confusion matrix based on classification. 
# Note this takes into account all groups and treatment and all values are log transformed through 
# cyt.plsda function. 
data.df <- cytodata

cyt.plsda(data.df[,-c(1,4)], title = "Example PLS-DA Analysis.pdf",
          colors = c("black", "purple", "red2"),
          bg = TRUE, scale = "log2",
          conf.mat = TRUE, var.num = 25,
          cv.opt = "loocv",
          comp.num = 2, pch.values = c(16,4,3), style = NULL,
          group.col = "Group", trt.col = "Treatment")

# Filtering data for specific groups and treatment
filt.data <- filter(data.df, Group != "ND", Treatment != "Unstimulated")
cyt.plsda(filt.data[,-c(1,4)], title = "Example PLS-DA Analysis 2.pdf",
          colors = c("black", "purple"),
          bg = TRUE, scale = "log2",
          conf.mat = TRUE, var.num = 25,
          cv.opt = "Mfold", fold.num = 5,
          comp.num = 3, pch.values = c(3,4), style = "3d",
          group.col = "Group", trt.col = "Treatment")
```

## 7. Principal Component Analysis (PCA)

```{r, echo=TRUE, warning=FALSE, message=FALSE}
data <- cytodata[,-c(1,4)]
data.df <- filter(data, Group != "ND" & Treatment != "Unstimulated")
data.df <- data.df[,-22]
cyt.pca(data.df, title = "Example PCA Analysis.pdf" ,colors = c("black", "red2"), scale = "log2", comp.num = 3, pch.values = c(16,4), style = "3D")
```

## 8. Volcano Plot

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating Volcano Plot
data.df <- cytodata[,-4]
volc_plot <- cyt.volc(data.df, "Group", cond1 = "T2D", cond2 = "ND", fold_change_thresh = 2.0, top_labels= 15)
ggsave("VolcanoPlot.png", plot = volc_plot$`T2D vs ND`, dpi = 300)

# Printing table (This is usually printed by default when the function is called and not saved as an object.)
print(volc_plot$`T2D vs ND`$data)

```

## 9. Heatmap

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating Heat map
cyt.heatmap(data = data.df,
                    scale = "log2",        # Optional scaling
                    annotation_col_name = "Group",
                    title = "Heatmap.png")
```

## 10. Dual Flashlight Plot

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Generating dual flashlights plot
data.df <- cytodata[,-c(1,3:4)]

dfp <- cyt.dualflashplot(data.df, group_var = "Group", group1 = "T2D", group2 = "ND", 
                  ssmd_thresh = -0.2, log2fc_thresh = 1, top_labels = 10)
ggsave("DualFlashlightPlot.png", plot = dfp$plot_env$p, dpi = 300, width = 3000, height = 2000, units = "px")

# Printing table (This is usually printed by default when the function is called and not saved as an object.)
print(dfp$data)

```

## 11. Machine Learning Models

### Using XGBoost for classification

```{r, echo=TRUE, warning=FALSE, message=FALSE}
data.df0 <- cytodata
data.df <- data.frame(data.df0[,1:4], log2(data.df0[,-c(1:4)]))
data.df <- data.df[,-c(1,3,4)]
data.df <- filter(data.df, Group != "ND")

xgb.results <- cyt.xgb(data = data.df, group_col = 'Group',
                   nrounds = 500, max_depth = 4, eta = 0.05,
                   nfold = 5, cv = TRUE, eval_metric = "mlogloss",
                   early_stopping_rounds = NULL, top_n_features = 10,
                   verbose = 0, plot_roc = TRUE)
```


### Using Random Forest for classification

```{r, echo=TRUE, warning=FALSE, message=FALSE}
rf.results <- cyt.rf(data = data.df, group_col = 'Group', k_folds = 5,
                  ntree = 1000, mtry = 4, run_rfcv = TRUE, plot_roc = TRUE)
```
