---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CytProfile

<!-- badges: start -->
<!-- badges: end -->

The goal of CytProfile is to conduct quality control using biological
    meaningful cutoff on raw measured values of cytokines. Test on distributional
    symmetry to suggest the adopt of transformation. Exploratory analysis including
    summary statistics, enriched boxplot, and barplots. Univariate analysis 
    and Multivariate analysis options are further available to dive deeper

## Installation

Before installation of the CytProfile package, make sure to install BiocManager and mix0mics packages using: 

```r
## install BiocManager
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager") 
## install mixOmics 
BiocManager::install('mixOmics')
```


You can install the development version of CytProfile from [GitHub](https://github.com/steal123/CytProfile) with:

``` r
# install.packages("devtools")
devtools::install_github("steal123/CytProfile")
```

## Example

This is a basic example which shows you how to analyze a data set:

```{r example}
library(CytProfile)
## basic example code
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.
