---
title: "Getting Started with CytoProfile"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CytoProfile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.keep = "none",
  fig.show = "hide"
)
```

# CytoProfile

The goal of CytoProfile is to conduct quality control using biological
meaningful cutoff on raw measured values of cytokines. Specifically, test on 
distributional symmetry to suggest the adopt of transformation. 
Conduct exploratory analysis including summary statistics, generate enriched 
barplots, and boxplots. Further, conduct univariate analysis and 
multivariate analysis for advance analysis. 

## Installation

Before installation of the CytoProfile package, make sure to install BiocManager and mixOmics packages using: 

```r
## install BiocManager
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager") 
## install mixOmics 
BiocManager::install('mixOmics')
```


You can install the development version of CytoProfile from [GitHub](https://github.com/saraswatsh/CytoProfile) with:

``` r
# install.packages("devtools")
devtools::install_github("saraswatsh/CytoProfile")
```
Install CytoProfile from CRAN with:

``` r
install.packages("CytoProfile")
```

## Example

Below are examples of using the functions provided in CytoProfile. Any saved or generated files that are PDF or PNG format will be found at in the [Output Folder](https://github.com/saraswatsh/CytoProfile/tree/main/output). 


## 1. Data Loading and set up

```{r, setup, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Loading all packages required
# Data manipulation and reshaping
library(dplyr)       # For data filtering, grouping, and summarising.
library(tidyr)       # For reshaping data (e.g., pivot_longer, pivot_wider).

# Plotting and visualization
library(ggplot2)     # For creating all the ggplot-based visualizations.
library(gridExtra)   # For arranging multiple plots on a single page.
library(ggrepel)     # For improved label placement in plots (e.g., volcano plots).
library(gplots)      # For heatmap.2, which is used to generate heatmaps.
library(plot3D)      # For creating 3D scatter plots in PCA and sPLS-DA analyses.
library(reshape2)    # For data transformation (e.g., melt) in cross-validation plots.

# Statistical analysis
library(mixOmics)    # For multivariate analyses (PCA, sPLS-DA, etc.).
library(e1071)     # For computing skewness and kurtosis.
library(pROC)        # For ROC curve generation in machine learning model evaluation.

# Machine learning
library(xgboost)     # For building XGBoost classification models.
library(randomForest) # For building Random Forest classification models.
library(caret)       # For cross-validation and other machine learning utilities.

# Package development and document rendering
library(knitr)       # For knitting RMarkdown files and setting chunk options.
library(devtools)    # For installing the development version of the package from GitHub.

# Load in the CytoProfile package
library(CytoProfile)

# Loading in data
data("ExampleData1")
data_df <- ExampleData1

# Creating a temporary directory to store output files
path <- tempdir(check = TRUE)

```

## 2. Exploratory Data Analysis

### Boxplots

```{r, EDA 1,  echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Generating boxplots to check for outliers for raw values
cyt_bp(data_df[, -c(1:3)], 
        pdf_title = file.path(path, "boxplot_by_cytokine_raw.pdf"))  
# Removing the first 3 columns to retain only continuous variables.

# Generating boxplots to check for outliers for log2 values
cyt_bp(data_df[, -c(1:3)], 
       pdf_title = file.path(path, "boxplot_by_cytokine_log2.pdf"),
       scale = "log2")  

```

### Group-Specific Boxplots

```{r, EDA 2, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Raw values for group-specific boxplots
cyt_bp2(data_df[, -c(3)], 
        pdf_title = file.path(path, "boxplot_by_group_and_treatment_raw.pdf"), 
        scale = NULL)

# Log2-transformed group-specific boxplots
cyt_bp2(data_df[, -c(3)], 
        pdf_title = file.path(path, "boxplot_by_group_and_treatment_log2.pdf"), 
        scale = "log2")


```

## 3. Skewness and Kurtosis 

```{r, EDA 3, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Histogram for overall raw data
cyt_skku(data_df[, -c(1:3)], 
         pdf_title = file.path(path, "skew_and_kurtosis.pdf"), 
         group_cols = NULL)

# Histogram with grouping (e.g., "Group")
cyt_skku(ExampleData1[, -c(2:3)], 
         pdf_title = file.path(path, "skew_and_kurtosis_2.pdf"), 
         group_cols = c("Group"))

```

## 4. Error Bar Plots

### Basic Error Bar Plot

```{r,EDA 4,  echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
cytokine_mat <- ExampleData1[, -c(1:3)]
cytokineNames <- colnames(cytokine_mat)
nCytokine <- length(cytokineNames)
results <- cyt_skku(ExampleData1[, -c(3)], print_res_log = TRUE, 
                    group_cols = c("Group", "Treatment"))

pdf(file = file.path(path, "bar_error_plot.pdf"))

oldpar <- par(no.readonly = TRUE)
par(mfrow = c(2,2), mar = c(4, 4, 2, 1))

for (k in 1:nCytokine) {
  center_df <- data.frame(name = rownames(results[,,k]), results[,,k])
  cyt_errbp(center_df,
            p_lab = FALSE, es_lab = FALSE, class_symbol = TRUE,
            y_lab = "Concentration in log2 scale", main = cytokineNames[k])
}
par(oldpar)
if (dev.cur() > 1) dev.off()

```

### Enriched Error Bar Plot with p-values and Effect Sizes

```{r, EDA 5,  echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
data_df <- ExampleData1[, -c(3)]
cyt_mat <- log2(data_df[, -c(1:2)])
data_df1 <- data.frame(data_df[, 1:2], cyt_mat)
cytokineNames <- colnames(cyt_mat)
nCytokine <- length(cytokineNames)
condt <- !is.na(cyt_mat) & (cyt_mat > 0)
Cutoff <- min(cyt_mat[condt], na.rm = TRUE) / 10

p_aov_mat <- matrix(NA, nrow = nCytokine, ncol = 3)
dimnames(p_aov_mat) <- list(cytokineNames, c("Group", "Treatment", "Interaction"))
p_groupComp_mat <- matrix(NA, nrow = nCytokine, ncol = 3)
dimnames(p_groupComp_mat) <- list(cytokineNames, c("2-1", "3-1", "3-2"))
ssmd_groupComp_stm_mat <- mD_groupComp_stm_mat <- p_groupComp_stm_mat <- p_groupComp_mat

for (i in 1:nCytokine) {
  Cytokine <- (cyt_mat[, i] + Cutoff)
  cytokine_aov <- aov(Cytokine ~ Group * Treatment, data = data_df)
  aov_table <- summary(cytokine_aov)[[1]]
  p_aov_mat[i, ] <- aov_table[1:3, 5]
  p_groupComp_mat[i, ] <- TukeyHSD(cytokine_aov)$Group[1:3, 4]
  p_groupComp_stm_mat[i, ] <- TukeyHSD(cytokine_aov)$`Group:Treatment`[1:3, 4]
  mD_groupComp_stm_mat[i, ] <- TukeyHSD(cytokine_aov)$`Group:Treatment`[1:3, 1]
  ssmd_groupComp_stm_mat[i, ] <- mD_groupComp_stm_mat[i, ] / sqrt(2 * aov_table["Residuals", "Mean Sq"])
}

results <- cyt_skku(ExampleData1[, -c(3)], print_res_log = TRUE, group_cols = c("Group", "Treatment"))

pdf(file = file.path(path, "bar_error_plot_enriched.pdf"))

oldpar <- par(no.readonly = TRUE)
par(mfrow = c(2,2), mar = c(3, 3, 2, 1))

for (k in 1:nCytokine) {
  result_mat <- results[1:9, , k]
  center_df <- data.frame(
    name = rownames(result_mat),
    result_mat[, c("center", "spread")],
    p.value = c(1, p_groupComp_stm_mat[k, 1:2]),
    effect.size = c(0, ssmd_groupComp_stm_mat[k, 1:2])
  )
  cyt_errbp(center_df, p_lab = TRUE, es_lab = TRUE, 
            class_symbol = TRUE, y_lab = "Concentration in log2 scale", 
            main = cytokineNames[k])
}
par(oldpar)
if (dev.cur() > 1) dev.off()

```

## 5. Univariate Analysis

### Two Sample T-test and Mann Whitney U Test

```{r, Univariate 1,  echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Performing Two Sample T-test and Mann Whitney U Test
data_df <- ExampleData1[, -c(3)]
data_df <- dplyr::filter(data_df, Group != "ND", Treatment != "Unstimulated")
# Two sample T-test
cyt_ttest(data_df[, c(1:2, 5:6)], scale = "log2", verbose = TRUE, format_output = TRUE)
# Mann-Whitney U Test
cyt_ttest(data_df[, c(1:2, 5:6)], verbose = TRUE)
```

### ANOVA Comparisons Test

```{r,Univariate 2,  echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Perform ANOVA comparisons test (example with 2 cytokines)
cyt_anova(data_df[, c(1:2, 5:6)], format_output = TRUE)
```

## 6. Multivariate Analysis

### Partial Least Squares Discriminant Analysis (PLS-DA) 

```{r, Multivariate 1, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# cyt_plsda function. 
data <- ExampleData1[, -c(3)]
data_df <- dplyr::filter(data, Group != "ND" & Treatment == "CD3/CD28")
cyt_splsda(data_df, 
           pdf_title = file.path(path, "example_spls_da_analysis.pdf"), 
           colors = c("black", "purple"),
           bg = FALSE, scale = "log2", ellipse = TRUE,
           conf_mat = FALSE, var_num = 25, 
           cv_opt = "loocv", comp_num = 2, 
           pch_values = c(16, 4), group_col = "Group", group_col2 = "Treatment", 
           roc = TRUE)
```

## 7. Principal Component Analysis (PCA)

```{r, Multivariate 2, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
data <- ExampleData1[, -c(3,23)]
data_df <- dplyr::filter(data, Group != "ND" & Treatment != "Unstimulated")
cyt_pca(data_df, 
        pdf_title = file.path(path, "example_pca_analysis.pdf"), 
        colors = c("black", "red2"), scale = "log2", 
        comp_num = 3, pch_values = c(16, 4), 
        style = "3D", group_col = "Group", group_col2 = "Treatment")
cyt_pca(data_df, 
        pdf_title = file.path(path, "example_pca_analysis_2.pdf"), 
        colors = c("black", "red2"), scale = "log2", 
        comp_num = 2, pch_values = c(16, 4), group_col = "Group")

```

## 8. Volcano Plot

```{r, EDA 6, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Generating Volcano Plot
data_df <- ExampleData1[, -c(2:3)]
cyt_volc(data_df, group_col = "Group", 
                      cond1 = "T2D", cond2 = "ND", 
                      fold_change_thresh = 2.0, 
                      top_labels = 15)

```

## 9. Heatmap

```{r,EDA 7, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE, fig.width=5, fig.height=4}
# Generating Heat map
cyt_heatmap(data = data_df,
            scale = "log2",        # Optional scaling
            annotation_col_name = "Group",
            title = NULL)
```

## 10. Dual Flashlight Plot

```{r,EDA 8, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Generating dual flashlights plot
data_df <- ExampleData1[, -c(2:3)]
dfp <- cyt_dualflashplot(data_df, group_var = "Group", 
                         group1 = "T2D", group2 = "ND", 
                         ssmd_thresh = -0.2, log2fc_thresh = 1, 
                         top_labels = 10)
# Print the plot
dfp
# Print the table data used for plotting
print(dfp$data)
```

## 11. Machine Learning Models

### Using XGBoost for classification

```{r, ML1, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Using XGBoost for classification
data_df0 <- ExampleData1
data_df <- data.frame(data_df0[, 1:3], log2(data_df0[, -c(1:3)]))
data_df <- data_df[, -c(2:3)]
data_df <- dplyr::filter(data_df, Group != "ND")

cyt_xgb(data = data_df, group_col = "Group",
                       nrounds = 500, max_depth = 4, eta = 0.05,
                       nfold = 5, cv = TRUE, eval_metric = "mlogloss",
                       early_stopping_rounds = NULL, top_n_features = 10,
                       verbose = 0, plot_roc = TRUE, print_results = FALSE)
```


### Using Random Forest for classification

```{r,ML2, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
# Using Random Forest for classification
cyt_rf(data = data_df, group_col = "Group", k_folds = 5,
                     ntree = 1000, mtry = 4, run_rfcv = TRUE,
                     plot_roc = TRUE, verbose = FALSE)
```
